DATA 621---Business Analytics and Data Mining

### 2020-09-17 HW1/HW1_draft_kellogg_AA.Rmd
Copyedit Model 1 & Model 2 and redo Model 2 so that it is consistent with the changed data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadData, include=FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
library(corrplot)
library(data.table)
library(tidymodels)
library(mice)
library(psych)
library(Hmisc)
library(MASS)
library(dplyr)
library(caret)
train_set <- read.csv('https://raw.githubusercontent.com/aadler/DT621_Fall2020_Group2/master/HW1/data/moneyball-training-data.csv')
eval_set <- read.csv('https://raw.githubusercontent.com/aadler/DT621_Fall2020_Group2/master/HW1/data/moneyball-evaluation-data.csv')
set.seed(9450)
```

## DATA EXPLORATION

### How often does the team win?

We are given a data set of 2,276 records containing 15 seasonal statistics and the
total number of wins a team had in a given year. On average, about 50% of games played 
are won (81 games out of 162), with the best individual season having 146 wins and the worst 
season having 0 wins. The data is normally distributed and most years have between 49 
and 112 wins (blue lines below). The nature of the distribution means there aren't too 
many extreme seasons where wins are significantly higher or lower than usual. This serves 
as a good gut-check for our final predictions; if the predicted wins are too high or too 
low, we know something in our model is probably off.

```{r wins, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 2.5, fig.width = 5, fig.align = 'center'}
avgWins <- mean(train_set$TARGET_WINS)
sdev <- sd(train_set$TARGET_WINS)
l1 <- mean(train_set$TARGET_WINS) - (2*sdev)
l2 <- mean(train_set$TARGET_WINS) + (2*sdev)
# histogram of wins
p <- ggplot(train_set, aes(x=TARGET_WINS)) + 
  geom_histogram() + 
  geom_vline(aes(xintercept= avgWins),
            color="red", linetype="dashed", size=1)+
  geom_vline(aes(xintercept=l1),
            color="blue", linetype="dashed", size=1) +
    geom_vline(aes(xintercept=l2),
            color="blue", linetype="dashed", size=1) +
  labs(title="Distribution of Wins",x="Number of wins", y = "Number of seasons")
p
```

### What's missing? 
A first look at the data shows that only about 8% of the records have a full set of 
information. The good news is that most of the missing values come from statistics 
that don't happen too often: hit-by-pitch (`TEAM_BATTING_HBP`, 92% missing!), caught 
stealing (`TEAM_BASERUN_CS`, 34% missing), and double plays (`TEAM_FIELDING_DP`, 13% 
missing). Since we have so little hit-by-pitch data, we expect that it doesn't contribute 
much to overall wins and will eliminate it from a few of the models we propose. The 
other two stats have less than half of the data missing, so we'll need to think of a 
clever way to fill in these values. The remaining missing information is from a 
combination of stolen bases and strikeouts by pitchers and batters (`TEAM_BASERUN_SB`,
`TEAM_PITCHING_SO`, `TEAM_BATTING_SO`). **It seems completely unreasonable** to have zero 
strike outs in a season, so this is something we'll most certainly have to impute.

```{r pctNull, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 2, fig.width = 5, fig.align = 'center'}
pct_null <- data.frame(do.call("rbind", map(train_set %>% dplyr::select(-INDEX), ~mean(is.na(.)))))
colnames(pct_null) <- c('PCT_NULL')
totalNulls <- pct_null %>%
  mutate(VARIABLE = rownames(.)) %>%
  arrange(desc(PCT_NULL)) %>%
  filter(PCT_NULL > 0) %>%
  dplyr::select(VARIABLE, PCT_NULL)
ggplot(totalNulls, aes(x=reorder(VARIABLE, PCT_NULL), y=PCT_NULL, label = round(PCT_NULL, 2))) + 
  geom_text(vjust = 0.5, hjust = -0.05)+
  geom_bar(stat = "identity") +
  ggtitle("Variables with Missing Information") +
  xlab("Statistic") + ylab("Percent Missing") + 
  coord_flip() +
  expand_limits(y = 1)
```

### Do the individual stats affect winning?

**Stats with an expected negative impact: ** Intuitively, we expect that Caught stealing, 
Errors, Hits allowed, Home Runs allowed, Strikeouts by batters, and Walks allowed would 
all have a **negative** impact on the total wins. In other words, as these values increase, 
we expect that the team is less likely to win.

```{r negFactors, echo = FALSE, message = FALSE, warning = FALSE}
negSet <- train_set %>%
  dplyr::select(TARGET_WINS, 
         'Errors' = TEAM_FIELDING_E, 
         'Hits allowed' = TEAM_PITCHING_H, 
         'Strikeouts by batters' = TEAM_BATTING_SO, 
         'Caught stealing' = TEAM_BASERUN_CS, 
         'Walks allowed' = TEAM_PITCHING_BB, 
         'Home Runs allowed' = TEAM_PITCHING_HR)
ggplot(data = negSet %>%
  gather(-TARGET_WINS, key = "STAT", value = "VALUE"), aes(x = VALUE, y = TARGET_WINS)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title="Number of wins with respect to recorded stats",x="", y = "Number of Wins") +
  facet_wrap(~ STAT, scales = "free")
```

When we take a closer look at the data, these negative relationships aren't obvious. In 
fact, only **Errors** and **Hits allowed** seem to have a negative impact on wins. **Caught**
**stealing** and **Strikeouts by batters** appear to be random; this means that whether the 
stat for a particular season is high or low doesn't affect the overall number of wins.

Even more interestingly, **Home Runs allowed** and **Walks allowed** have the *opposite* 
effect; as these stats increase, so do the number of wins!

**Stats with an expected positive impact: ** We can look at the same information for the 
stats that we expect to have a **positive** effect on wins: Base hits, Doubles, Triples, Home 
Runs, Walks, Batters getting hit by pitches, Stolen bases, Double Plays, and Strikeouts by 
pitchers.

```{r posFactors, echo = FALSE, message = FALSE, warning = FALSE}
posSet <- train_set %>%
  dplyr::select(TARGET_WINS, 
         'Base hits' = TEAM_BATTING_H, 
         'Doubles' = TEAM_BATTING_2B, 
         'Triples' = TEAM_BATTING_3B, 
         'Home Runs' = TEAM_BATTING_HR, 
         'Walks' = TEAM_BATTING_BB, 
         'Batters hit' = TEAM_BATTING_HBP,
         'Stolen bases' = TEAM_BASERUN_SB,
         'Double plays' = TEAM_FIELDING_DP,
         'Strikeouts' = TEAM_PITCHING_SO)
ggplot(data = posSet %>%
  gather(-TARGET_WINS, key = "STAT", value = "VALUE"), aes(x = VALUE, y = TARGET_WINS)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title="Number of Wins with Respect to Recorded Stats",x="", y = "Number of Wins") +
  facet_wrap(~ STAT, scales = "free")
```

Many of these stats *do* seem to have an effect on the number of wins, most notably,
**Base hits** and **Walks**. We see weaker positive relationships for **Home runs**, 
**Doubles**, **Triples**, and **Stolen bases**. This makes sense when we think about it; 
these things tend to happen less often in games than pure base hits and walks, so they don't 
have as much of an effect on winning. Finally, **Double plays** and **Batters hit** don't 
appear to have any correlation with the number of wins. Once again, this intuitively makes 
sense because they are less likely to happen in a game.

@@

### Are some stats more skewed than others?

Before using any of the statistics in a model, we need to take a closer look at the variation 
in the data. We call out-of-the-ordinary values (exceptionally high or low values) **outliers**. 
We need to take these into account in our modeling because we want to make sure our predictions 
aren't skewed because of them.

```{r summaryInfo, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3, fig.width = 5, fig.align = 'center'}
#summary(train_set)
totalOutliers <- data.frame(sapply(train_set %>% dplyr::select(-INDEX, -TARGET_WINS, - TEAM_BATTING_HBP), 
       function(x){length(boxplot.stats(x)$out)/nrow(train_set)}))
totalOutliers$VARIABLE_NM <- rownames(totalOutliers)
colnames(totalOutliers) <- c('PCT_OUTLIERS', 'VARIABLE_NM')
ggplot(totalOutliers , aes(x=reorder(VARIABLE_NM, PCT_OUTLIERS), y=PCT_OUTLIERS, label = round(PCT_OUTLIERS, 3))) + 
  geom_text(vjust = 0.5, hjust = -0.05)+
  geom_bar(stat = "identity") +
  ggtitle("Percentage of Outliers") +
  xlab("Statistic") + ylab("Percent of Data that is an Outlier") + 
  coord_flip() +
  expand_limits(y = 0.15)
```

As we can see, some of the provided statistics are well-balanced in the sense that there are very 
*few* (or no) extreme values. **Home Runs allowed** (`TEAM_PITCHING_HR`), **Strikeouts by batters** 
(`TEAM_BATTING_SO`), and **Home Runs by batters** (`TEAM_BATTING_HR`) are examples of this.

```{r fewOutliers, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 2, fig.width = 6, fig.align = 'center'}
set1 <-  train_set %>%
  dplyr::select('Home Runs by batters' = TEAM_BATTING_HR, 
         'Strikeouts by batters' = TEAM_BATTING_SO,  
         'Home Runs allowed' = TEAM_PITCHING_HR) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)
vals <- ggplot(set1 , aes(x=stat, y=value)) + 
    geom_boxplot() +
  labs(title="",x="", y = "Value") +
  facet_wrap(~stat, scale="free")
vals
```
Some things to note about each of these statistics:

* **Home Runs allowed** (average = ~100/year) and **Home Runs by batters** (average = ~106/year) 
have a very similar mid-range distribution (50% of the data lies between ~50 and 150). 
The slight difference in average stats means that teams tend to have a higher number
of Home Runs than the opposition team. 
* The only thing that stands out about **Strikeouts by batters** (average = ~736/year) is 
how nearly perfectly normal it is. 50% of the data is between about 500 and 1000 and 
there are absolutely no outliers in the dataset! This means that there were no surprisingly
high or low seasons.

Conversely, some of the stats have a very *high* number of outliers, indicating that there
are some seasons with some abnormally high or low values. **Errors** (`TEAM_FIELDING_E`), 
**Hits allowed** (`TEAM_PITCHING_H`), **Walks by batters** (`TEAM_BATTING_BB`), and 
**Stolen bases** (`TEAM_BASERUN_SB`) are examples of this.

```{r manyOutliers, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 2, fig.width = 6, fig.align = 'center'}
set2 <-  train_set %>%
  dplyr::select('Walks by batters' = TEAM_BATTING_BB, 
         'Stolen bases' = TEAM_BASERUN_SB,  
         'Hits allowed' = TEAM_PITCHING_H,
         'Errors' = TEAM_FIELDING_E) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)
vals2 <- ggplot(set2 , aes(x=stat, y=value)) + 
    geom_boxplot() +
  labs(title="",x="", y = "Value") +
  facet_wrap(~ stat, scale = 'free', ncol = 4)
vals2
```
Some things to note about each of these statistics:

* All of the outliers for **Errors**, **Hits allowed**, and **Stolen bases** are above the upper 
tail of the data set. This is further illustrated by the mean and median values for these stats; in 
all instances, the mean per year (Errors = ~246/year, Hits allowed = ~1779/year, Stolen bases = ~125/year) 
are higher than the median per year (Errors = ~159/year, Hits allowed = ~1518/year, Stolen bases = 
~101/year). This means that some seasons with exceptionally high values skew the dataset. 
* There are a few *very* extreme outliers for **Hits allowed**. The maximum value is 30,132, which
is over 16 times the average number of hits allowed per season!
* There are outliers both above *and* below the tails of the data for the **Walks by batters** stat. 
This means that we have exceptionally low (min = 0!) and exceptionally high (max = 878) seasons.

The remaining stats, **Walks allowed** (`TEAM_PITCHING_BB`), **Base Hits by batters** (`TEAM_BATTING_H`), 
**Caught stealing** (`TEAM_BASERUN_CS`), **Strikeouts by pitchers** (`TEAM_PITCHING_SO`), **Double plays** 
(`TEAM_FIELDING_DP`), **Triples by batters** (`TEAM_BATTING_3B`), and **Doubles by batters** 
(`TEAM_BATTING_2B`) have between 29 and 99 outliers.

```{r otherOutliers, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3.5, fig.width = 7, fig.align = 'center'}
set3 <-  train_set %>%
  dplyr::select('Base Hits by batters' = TEAM_BATTING_H, 
         'Doubles by batters' = TEAM_BATTING_2B,  
         'Triples by batters' = TEAM_BATTING_3B,
         'Caught stealing' = TEAM_BASERUN_CS,
         'Walks allowed' = TEAM_PITCHING_BB,
         'Strikeouts by pitchers' = TEAM_PITCHING_SO,
         'Double plays' = TEAM_FIELDING_DP) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)
vals3 <- ggplot(set3 , aes(x=stat, y=value)) + 
    geom_boxplot() +
  labs(title="",x="", y = "Value") +
  facet_wrap(~ stat, scale = 'free', ncol = 4)
vals3
```

@@

* All variables are very narrowly distributed, meaning that most of the data falls within
a small range. 
* **Strikeouts by pitchers** and **Walks allowed** have a few very extreme outliers; these 
represent seasons that have abnormally high values for the statistics. 
* The average number of pure **Base Hits** (1469/season) is greater than the average number
of Doubles (~241/season) and Triples (~55/season). This isn't at all surprising, but serves 
as a good gut check on the validity of the data.

### Are stats correlated?

We would expect that a few things in the dataset might be correlated: perhaps number of errors and
hits/homeruns allowed or the number of base hits by batters and doubles/triples/homeruns. We can
visualize the correlations between the statistics to determine if there is a significant relationship
between the them: blue dots represent positively correlated variables (as one increases,
so does the other) and red dots represent negatively correlated variables (as one increases,
the other decreases).

@@

* **Errors** are highly, negatively correlated with walks by batters, strikeouts by batters, and
homeruns (both by batters and allowed). 
* **Triples by batters** are highly, negatively correlated with strikeouts by batters and homeruns
(both by batters and allowed).
* **Strikeouts by batters** are highly, positively correlated with homeruns (both by batters and
allowed). 
* **Homeruns by batters** and **Homeruns allowed** are both positively correlated with walks by 
batters.
* As expected, **basehits** are positively correlated with doubles and triples.

@@

## DATA PREPARATION

Now we have a good idea of the data we are looking at we can take the next steps to 
prepare it for building a solid model.

### Outlier Removal

As we saw in the data analysis, there are some outlier concerns for some of the variables,
so we will need to account for this in our modeling. When performing the processes of outlier 
removal, a cautious approach is always best. Each outlier is evaluated to ensure:

@@

```{r outliers_1, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 8, fig.align = 'center'}
train_set %>%
  dplyr::select(-TARGET_WINS, -INDEX) %>%
  pivot_longer(everything(), 
               names_to ="Variable", 
               values_to="Value") %>%
  ggplot(aes(x=Variable, y=Value)) +
  geom_boxplot(na.rm = TRUE) + 
  labs(title="Outliers for each Statistic",x="Variable", y = "Value") +
  coord_flip()
```

Off the bat, there are some clear issues with two of the variables: `TEAM_PITCHING_SO` and
`TEAM_PITCHING_H.`

* **TEAM_PITCHING_SO**: The largest outlier is close to 20,000, which averages ~123 strikeouts
per game. The MLB team pitching strikeout record is 1,450 and set in 2014 (CLE). There are a 
minimum of 27 at-bats per game and most teams average ~35 at-bats, thus making 123 strikeouts
impossible. 
* **TEAM_PITCHING_H**: The largest outlier for this vaiable is over 30,000 which is highly 
unlikely.

@@

```{r iqr_removal_SO, include=FALSE}
train_set<- train_set %>%
  na.omit() %>%
  summarise(iqr = IQR(TEAM_PITCHING_SO)) %>%
  bind_cols(
            train_set 
            )  %>%
  filter(
    TEAM_PITCHING_SO > quantile(TEAM_PITCHING_SO,
                                probs=c(0.25), 
                                na.rm = TRUE) - 1.5*iqr, 
    TEAM_PITCHING_SO < quantile(TEAM_PITCHING_SO, 
                                probs=c(0.75), 
                                na.rm=TRUE) + 1.5*iqr
        ) %>%
  dplyr::select(-iqr)
```

```{r iqr_removal_H, include=FALSE}
train_set<- train_set %>%
  na.omit() %>%
  summarise(iqr = IQR(TEAM_PITCHING_H)) %>%
  bind_cols(
            train_set 
            )  %>%
  filter(
    TEAM_PITCHING_H > quantile(TEAM_PITCHING_H,
                                probs=c(0.25), 
                                na.rm = TRUE) - 1.5*iqr, 
    TEAM_PITCHING_H < quantile(TEAM_PITCHING_H, 
                                probs=c(0.75), 
                                na.rm=TRUE) + 1.5*iqr
        ) %>%
  dplyr::select(-iqr)
```

We can take a look at the distributions of these variables after outlier removal:

```{r outlier_boxplot, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 8, fig.align = 'center'}
train_set %>%
  dplyr::select(-TARGET_WINS, -INDEX) %>%
  pivot_longer(everything(), 
               names_to ="Variable", 
               values_to="Value") %>%
  ggplot(aes(x=Variable, y=Value)) +
  geom_boxplot(na.rm = TRUE) + 
  labs(title="Outliers for each Statistic",x="Variable", y = "Value") +
  coord_flip()
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, error=FALSE, warning=FALSE, message=FALSE)
```

```{r loadData, include=FALSE}
library(tidyverse)
library(ggplot2)
library(corrplot)
library(tidymodels)
library(mice)
library(psych)
library(Hmisc)
library(MASS)
library(dplyr)
library(caret)
library(data.table)
train_set <- read.csv('https://raw.githubusercontent.com/aadler/DT621_Fall2020_Group2/master/HW1/data/moneyball-training-data.csv')
eval_set <- read.csv('https://raw.githubusercontent.com/aadler/DT621_Fall2020_Group2/master/HW1/data/moneyball-evaluation-data.csv')
ts2 <- train_set
es2 <- eval_set
set.seed(9450)
```

# DATA EXPLORATION
## How Often Does The Team Win?
We are given a data set of 2,276 records containing 15 seasonal statistics and
the total number of wins a team had in a given year. On average, about 50% of
games played are won (81 games out of 162), with the best individual season
having 146 wins and the worst season having 0 wins. The data is normally
distributed and most years have between 49 and 112 wins (blue lines below). The
nature of the distribution means there aren't too many extreme seasons where
wins are significantly higher or lower than usual. This serves as a good
gut-check for our final predictions; if the predicted wins are too high or too 
low, we know something in our model is probably off.

```{r wins, fig.height = 2.5, fig.width = 5, fig.align = 'center'}
avgWins <- mean(train_set$TARGET_WINS)
sdev <- sd(train_set$TARGET_WINS)
l1 <- mean(train_set$TARGET_WINS) - (2 * sdev)
l2 <- mean(train_set$TARGET_WINS) + (2 * sdev)

# histogram of wins

p <- ggplot(train_set, aes(x=TARGET_WINS)) +
  geom_histogram() + 
  geom_vline(aes(xintercept = avgWins),
            color = "red", linetype = "dashed", size = 1)+
  geom_vline(aes(xintercept = l1),
            color="blue", linetype = "dashed", size = 1) +
    geom_vline(aes(xintercept = l2),
            color = "blue", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Wins", x = "Number of wins",
       y = "Number of seasons")
p
```

## What's Missing? 
A first look at the data shows that only about 8% of the records have a full set
of information. The good news is that most of the missing values come from
statistics that don't happen too often: hit-by-pitch (`TEAM_BATTING_HBP`, 92%
missing!), caught stealing (`TEAM_BASERUN_CS`, 34% missing), and double plays
(`TEAM_FIELDING_DP`, 13% missing). Since we have so little hit-by-pitch data, we
expect that it doesn't contribute much to overall wins and will eliminate it
from a few of the models we propose. The other two stats have less than half of
the data missing, so we'll need to think of a clever way to fill in these
values. The remaining missing information is from a combination of stolen bases
and strikeouts by pitchers and batters (`TEAM_BASERUN_SB`, `TEAM_PITCHING_SO`,
`TEAM_BATTING_SO`). **It seems completely unreasonable** to have zero strike
outs in a season, so this is something we'll most certainly have to impute.

```{r pctNull, fig.height = 2, fig.width = 5, fig.align = 'center'}
pct_null <- data.frame(do.call("rbind", map(train_set %>% dplyr::select(-INDEX),
                                            ~ mean(is.na(.)))))
colnames(pct_null) <- c('PCT_NULL')
totalNulls <- pct_null %>%
  mutate(VARIABLE = rownames(.)) %>%
  arrange(desc(PCT_NULL)) %>%
  filter(PCT_NULL > 0) %>%
  dplyr::select(VARIABLE, PCT_NULL)
ggplot(totalNulls, aes(x = reorder(VARIABLE, PCT_NULL), y = PCT_NULL,
                       label = round(PCT_NULL, 2))) + 
  geom_text(vjust = 0.5, hjust = -0.05)+
  geom_bar(stat = "identity") +
  ggtitle("Variables with Missing Information") +
  xlab("Statistic") + ylab("Percent Missing") + 
  coord_flip() + expand_limits(y = 1)
```

## Do the Individual Stats Affect Winning?
**Stats with an expected negative impact:** Intuitively, we expect that Caught
Stealing, Errors, Hits Allowed, Home Runs Allowed, Strikeouts by Batters, and
Walks Allowed would all have a **negative** impact on the total wins. In other
words, as these values increase, we expect that the team is less likely to win.

```{r negFactors}
negSet <- train_set %>%
  dplyr::select(TARGET_WINS, 
         'Errors' = TEAM_FIELDING_E, 
         'Hits allowed' = TEAM_PITCHING_H, 
         'Strikeouts by batters' = TEAM_BATTING_SO, 
         'Caught stealing' = TEAM_BASERUN_CS, 
         'Walks allowed' = TEAM_PITCHING_BB, 
         'Home Runs allowed' = TEAM_PITCHING_HR)
ggplot(data = negSet %>%
  gather(-TARGET_WINS, key = "STAT", value = "VALUE"),
  aes(x = VALUE, y = TARGET_WINS)) + 
  geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Number of wins with respect to recorded stats",
       x = "", y = "Number of Wins") + facet_wrap(~ STAT, scales = "free")
```

When we take a closer look at the data, these negative relationships aren't
obvious. In fact, only **Errors** and **Hits Allowed** seem to have a negative
impact on wins. **Caught Stealing** and **Strikeouts by Batters** appear to be
random; this means that whether the stat for a particular season is high or low
doesn't affect the overall number of wins.

Even more interestingly, **Home Runs Allowed** and **Walks Allowed** have the
*opposite* effect---as these stats increase, so do the number of wins!

**Stats with an expected positive impact:** We can look at the same information
for the stats that we expect to have a **positive** effect on wins: Base Hits,
Doubles, Triples, Home Runs, Walks, Batters getting hit by pitches, Stolen
Bases, Double Plays, and Strikeouts by Pitchers.

```{r posFactors}
posSet <- train_set %>%
  dplyr::select(TARGET_WINS, 
         'Base hits' = TEAM_BATTING_H, 
         'Doubles' = TEAM_BATTING_2B, 
         'Triples' = TEAM_BATTING_3B, 
         'Home Runs' = TEAM_BATTING_HR, 
         'Walks' = TEAM_BATTING_BB, 
         'Batters hit' = TEAM_BATTING_HBP,
         'Stolen bases' = TEAM_BASERUN_SB,
         'Double plays' = TEAM_FIELDING_DP,
         'Strikeouts' = TEAM_PITCHING_SO)
ggplot(data = posSet %>%
  gather(-TARGET_WINS, key = "STAT", value = "VALUE"),
  aes(x = VALUE, y = TARGET_WINS)) + geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Number of Wins with Respect to Recorded Stats", x = "",
       y = "Number of Wins") + facet_wrap(~ STAT, scales = "free")
```

Many of these stats *do* seem to have an effect on the number of wins, most
notably, **Base Hits** and **Walks**. We see weaker positive relationships for
**Home Runs**, **Doubles**, **Triples**, and **Stolen Bases**. This makes sense
when we think about it; these things tend to happen less often in games than
pure base hits and walks, so they don't have as much of an effect on winning.
Finally, **Double Plays** and **Batters Hit** don't appear to have any
correlation with the number of wins. Once again, this intuitively makes sense
because they are less likely to happen in a game.

@@

## Are Some Stats More Skewed Than Others?
Before using any of the statistics in a model, we need to take a closer look at
the variation in the data. We call out-of-the-ordinary values (exceptionally
high or low values) **outliers**. We need to take these into account in our
modeling because we want to make sure our predictions aren't skewed because of
them.

```{r summaryInfo, fig.height = 3, fig.width = 5, fig.align = 'center'}
#summary(train_set)
totalOutliers <- data.frame(
  sapply(train_set %>% dplyr::select(-INDEX, -TARGET_WINS, - TEAM_BATTING_HBP), 
       function(x){length(boxplot.stats(x)$out)/nrow(train_set)}))
totalOutliers$VARIABLE_NM <- rownames(totalOutliers)
colnames(totalOutliers) <- c('PCT_OUTLIERS', 'VARIABLE_NM')
ggplot(totalOutliers,
       aes(x = reorder(VARIABLE_NM, PCT_OUTLIERS), y=PCT_OUTLIERS,
           label = round(PCT_OUTLIERS, 3))) + 
  geom_text(vjust = 0.5, hjust = -0.05)+ geom_bar(stat = "identity") +
  ggtitle("Percentage of Outliers") + xlab("Statistic") +
  ylab("Percent of Data that is an Outlier") + coord_flip() +
  expand_limits(y = 0.15)
```

As we can see, some of the provided statistics are well-balanced in the sense
that there are very *few* (or no) extreme values. **Home Runs allowed**
(`TEAM_PITCHING_HR`), **Strikeouts by batters** (`TEAM_BATTING_SO`), and
**Home Runs by batters** (`TEAM_BATTING_HR`) are some examples.

```{r fewOutliers, fig.height = 2, fig.width = 6, fig.align = 'center'}
set1 <- train_set %>%
  dplyr::select('Home Runs by batters' = TEAM_BATTING_HR, 
         'Strikeouts by batters' = TEAM_BATTING_SO,  
         'Home Runs allowed' = TEAM_PITCHING_HR) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)
vals <- ggplot(set1 , aes(x = stat, y = value)) + 
    geom_boxplot() + labs(title = "", x = "", y = "Value") +
  facet_wrap( ~ stat, scales = "free")
vals
```

Some things to note about each of these statistics:

  * **Home Runs Allowed** (average ~100/year) and **Home Runs by Batters**
  (average ~106/year) have a very similar mid-range distribution (50% of the
  data lies between ~50 and 150). The slight difference in average stats means
  that teams tend to have a higher number of Home Runs than the opposition team.
  * The only thing that stands out about **Strikeouts by Batters** (average
  ~736/year) is how nearly perfectly normal it is. 50% of the data is between
  about 500 and 1000 and there are absolutely no outliers in the dataset! This
  means that there were no surprisingly high or low seasons.

Conversely, some of the stats have a very *high* number of outliers, indicating
that there are some seasons with some abnormally high or low values. **Errors**
(`TEAM_FIELDING_E`), **Hits allowed** (`TEAM_PITCHING_H`), **Walks by batters**
(`TEAM_BATTING_BB`), and **Stolen bases** (`TEAM_BASERUN_SB`) are some examples.

```{r manyOutliers, fig.height = 2, fig.width = 6, fig.align = 'center'}
set2 <-  train_set %>%
  dplyr::select('Walks by batters' = TEAM_BATTING_BB, 
         'Stolen bases' = TEAM_BASERUN_SB,  
         'Hits allowed' = TEAM_PITCHING_H,
         'Errors' = TEAM_FIELDING_E) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)
vals2 <- ggplot(set2 , aes(x=stat, y=value)) + 
    geom_boxplot() +
  labs(title = "",x = "", y = "Value") +
  facet_wrap(~ stat, scales = 'free', ncol = 4)
vals2
```

Some things to note about each of these statistics:

  * All of the outliers for **Errors**, **Hits Allowed**, and **Stolen Bases**
  are above the upper tail of the data set. This is further illustrated by the
  mean and median values for these stats; in all instances, the means per year
  (Errors ~246/year, Hits Allowed ~1779/year, Stolen Bases ~125/year) are higher
  than the medians per year (Errors ~159/year, Hits Allowed ~1518/year, Stolen
  Bases ~101/year). This means that some seasons with exceptionally high values
  skew the dataset.
  * There are a few *very* extreme outliers for **Hits Allowed**. The maximum
  value is 30,132, which is over 16 times the average number of hits allowed per
  season!
  * There are outliers both above *and* below the tails of the data for the
  **Walks by Batters** stat. This means that we have exceptionally low
  (min = 0!) and exceptionally high (max = 878) seasons.

The remaining stats, **Walks Allowed** (`TEAM_PITCHING_BB`),
**Base Hits by Batters** (`TEAM_BATTING_H`), **Caught Stealing**
(`TEAM_BASERUN_CS`), **Strikeouts by Pitchers** (`TEAM_PITCHING_SO`),
**Double Plays** (`TEAM_FIELDING_DP`), **Triples by Batters**
(`TEAM_BATTING_3B`), and **Doubles by Batters** (`TEAM_BATTING_2B`) have between
29 and 99 outliers.

```{r otherOutliers, fig.height = 3.5, fig.width = 7, fig.align = 'center'}
set3 <-  train_set %>%
  dplyr::select('Base Hits by batters' = TEAM_BATTING_H, 
         'Doubles by batters' = TEAM_BATTING_2B,  
         'Triples by batters' = TEAM_BATTING_3B,
         'Caught stealing' = TEAM_BASERUN_CS,
         'Walks allowed' = TEAM_PITCHING_BB,
         'Strikeouts by pitchers' = TEAM_PITCHING_SO,
         'Double plays' = TEAM_FIELDING_DP) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)
vals3 <- ggplot(set3, aes(x=stat, y=value)) + 
    geom_boxplot() +
  labs(title = "", x = "",y = "Value") +
  facet_wrap(~ stat, scales = 'free', ncol = 4)
vals3
```

@@

  * All variables are very narrowly distributed, meaning that most of the data
  falls within a small range. 
  * **Strikeouts by Pitchers** and **Walks Allowed** have a few very extreme
  outliers; these   represent seasons that have abnormally high values for the
  statistics. 
  * The average number of pure **Base Hits** (1469/season) is greater than the
  average number of Doubles (~241/season) and Triples (~55/season). This isn't
  at all surprising, but serves as a good gut check on the validity of the data.

## Are Stats Correlated?
We would expect that a few things in the dataset might be correlated: perhaps
number of errors and hits/homeruns allowed or the number of base hits by batters
and doubles/triples/homeruns. We can visualize the correlations between the
statistics to determine if there is a significant relationship between the them:
blue dots represent positively correlated variables (as one increases, so does
the other) and red dots represent negatively correlated variables (as one
increases, the other decreases).

@@

  * **Errors** are highly, negatively correlated with walks by batters,
  strikeouts by batters, and homeruns (both by batters and allowed).
  * **Triples by Batters** are highly, negatively correlated with strikeouts by
  batters and homeruns (both by batters and allowed).
  * **Strikeouts by Batters** are highly, positively correlated with homeruns
  (both by batters and allowed). 
  * **Homeruns by Batters** and **Homeruns Allowed** are both positively
  correlated with walks by batters.
  * As expected, **Base Hits** are positively correlated with doubles and
  triples.

@@

# DATA PREPARATION
Now we have a good idea of the data we are looking at we can take the next steps
to prepare it for building a solid model.

## Outlier Removal
As we saw in the data analysis, there are some outlier concerns for some of the
variables, so we will need to account for this in our modeling. When performing
the processes of outlier removal, a cautious approach is always best. Each
outlier is evaluated to ensure:

@@

```{r outliers_1, fig.height = 4, fig.width = 8, fig.align = 'center'}
train_set %>%
  dplyr::select(-TARGET_WINS, -INDEX) %>%
  pivot_longer(everything(), 
               names_to = "Variable", 
               values_to = "Value") %>%
  ggplot(aes(x = Variable, y = Value)) +
  geom_boxplot(na.rm = TRUE) + 
  labs(title="Outliers for each Statistic",x="Variable", y = "Value") +
  coord_flip()
```

Off the bat (pun intended), there are some clear issues with two of the
variables: `TEAM_PITCHING_SO` and `TEAM_PITCHING_H.`

  * **TEAM_PITCHING_SO**: The largest outlier is close to 20,000, which averages
  ~123 strikeouts per game. This is only possible if every one of the 162 games
  went approximately 40 innings, and each out was a strikeout---which clearly has
  never happened.
  * **TEAM_PITCHING_H**: The largest outlier for this vaiable is over 30,000
  which is simlarly highly unlikely.

@@

```{r iqr_removal_SO}
train_set<- train_set %>%
  na.omit() %>%
  summarise(iqr = IQR(TEAM_PITCHING_SO)) %>%
  bind_cols(
            train_set 
            )  %>%
  filter(
    TEAM_PITCHING_SO > quantile(TEAM_PITCHING_SO,
                                probs = c(0.25), 
                                na.rm = TRUE) - 1.5 * iqr, 
    TEAM_PITCHING_SO < quantile(TEAM_PITCHING_SO, 
                                probs = c(0.75), 
                                na.rm = TRUE) + 1.5 * iqr
        ) %>%
  dplyr::select(-iqr)
```
```{r iqr_removal_H}
train_set<- train_set %>%
  na.omit() %>%
  summarise(iqr = IQR(TEAM_PITCHING_H)) %>%
  bind_cols(
            train_set 
            )  %>%
  filter(
    TEAM_PITCHING_H > quantile(TEAM_PITCHING_H,
                                probs = c(0.25), 
                                na.rm = TRUE) - 1.5 * iqr, 
    TEAM_PITCHING_H < quantile(TEAM_PITCHING_H, 
                                probs = c(0.75), 
                                na.rm = TRUE) + 1.5 * iqr
        ) %>%
  dplyr::select(-iqr)
```

We can take a look at the distributions of these variables after removing the
outliers:

```{r outlier_boxplot, fig.height = 4, fig.width = 8, fig.align = 'center'}
train_set %>%
  dplyr::select(-TARGET_WINS, -INDEX) %>%
  pivot_longer(everything(), 
               names_to ="Variable", 
               values_to="Value") %>%
  ggplot(aes(x=Variable, y=Value)) +
  geom_boxplot(na.rm = TRUE) + 
  labs(title="Outliers for each Statistic",x="Variable", y = "Value") +
  coord_flip()
```
